Wildcard: End user modification of web applications through a data table

### 2020-01-07 paper.tex
starting to add dnj edits

\author{Geoffrey Litt}
\authorinfo{is bla bla bla}
\author{Daniel Jackson}
\authorinfo{is bla bla bla}
\affiliation{Massachusetts Institute of Technology}

\keywords{programming journal, paper formatting, submission preparation} % please provide 1--5 keywords

%%%%%%%%%%%%%%%%%%
%% These data MUST be filled for your submission. (see 5.3)
\paperdetails{
  %% perspective options are: art, sciencetheoretical, scienceempirical, engineering.
  %% Choose exactly the one that best describes this work. (see 2.1)
  perspective=art,
  %% State one or more areas, separated by a comma. (see 2.2)
  %% Please see list of areas in http://programming-journal.org/cfp/
  %% The list is open-ended, so use other areas if yours is/are not listed.
  area={Social Coding, General-purpose programming},
  %% You may choose the license for your paper (see 3.)
  %% License options include: cc-by (default), cc-by-nc
  % license=cc-by,
}
%%%%%%%%%%%%%%%%%%

@@

% Please always include the abstract.
% The abstract MUST be written according to the directives stated in 
% http://programming-journal.org/submission/
% Failure to adhere to the abstract directives may result in the paper
% being returned to the authors.
\begin{abstract}
Browser extensions and user scripts can modify web applications in
useful ways, but many people have unique needs that aren't met by
existing extensions. Today, most of those people can't build their own
browser extensions, since they can't program in Javascript.

Wildcard is a platform that empowers anyone to build browser extensions
without doing traditional programming. Wildcard shows the main data from
a web page in a table that maintains a bidirectional connection to the
original page. By directly manipulating the table, people can
sort/filter content, add private annotations, and use spreadsheet
formulas to fetch data from other websites.
\end{abstract}

@@

In 2012, the travel site Airbnb removed the ability to sort listings by
price. Users could still filter down to a price range, but could no
longer view the cheapest listings first. Many users complained on online
message boards that the change seemed hostile to users. ``It's so
frustrating!..What is the logic behind not having this function?'' says
one user on the
\href{https://community.withairbnb.com/t5/Hosting/Sorting-listing-by-price/td-p/559404}{Airbnb
support forum}. Alas, the feature remains missing to this day.

This is a familiar situation in a world of web applications that are
frequently updated without user consent. When web software doesn't quite
meet users' needs, their only recourse is to complain to the developers
and hope someone listens. Sometimes there is a way to use a browser
extension or user script to fix the issue, but making an extension
requires sophisticated programming skills. We see this as a waste of the
openness of the Web platform, and of the pliability of software in
general. If users could make tweaks to web applications without
programming, a vast number of people would gain the agency to modify
software to meet their unique preferences.

In this paper, we introduce a system called Wildcard\footnote{Wildcard
  was the original internal name for Hypercard, which pioneered both
  user modifiable software and the modern Web.} that aims to meet this
need. Wildcard adds a panel to the bottom of a web page that shows a
structured table view of some of the main data in the page. The table
maintains a bidirectional connection to the original page---when the
user manipulates the table, the original page also gets modified, and
vice versa.

In Wildcard, sorting Airbnb listings by price takes just one click on a
table header, with no programming required by the end user. Beyond
sorting and filtering data, Wildcard also supports a variety of other
use cases like accessing third party APIs, performing small
computations, recording private user annotations, and using alternate UI
widgets. While Wildcard does not support all changes someone might want
to make to a website, the goal of the project is to make a broad subset
of changes easily accessible to end users.

Under the hood, there is no magic---Wildcard currently requires a
programmer to manually define an adapter between the original page and
the data table using web scraping techniques. While programming is
required for part of the process, this forms a useful collaboration
between the programmer and the end user, since the end user is able to
perform many different modifications once an adapter is defined. We also
plan to eventually reduce the burden of making adapters (discussed in
more detail later).

\hypertarget{demo-booking-a-trip-with-wildcard}{%
\section{Demo: Booking a trip with
Wildcard}\label{demo-booking-a-trip-with-wildcard}}

To get a sense of the user experience of Wildcard, let's see an example
of how an end user can modify a website with Wildcard in practice.

For example, in Fig.~\ref{fig:table} we open up a table view that
corresponds to search results on the Airbnb travel site.

\begin{figure}
\hypertarget{fig:table}{%
\centering
\includegraphics{media/opentable.png}
\caption{Opening a table corresponding to search results on
Airbnb}\label{fig:table}
}
\end{figure}

\begin{itemize}
\tightlist
\item
  Sorting search results:

  \begin{itemize}
  \tightlist
  \item
    Airbnb took away search
  \item
    once you see the tabular view, there's an obvious interaction
    available to anyone familiar with tables
  \end{itemize}
\item
  Injecting new data into the page

  \begin{itemize}
  \tightlist
  \item
    Take own custom notes, saved in the browser. Maybe shared in the
    future
  \end{itemize}
\item
  Formulas

  \begin{itemize}
  \tightlist
  \item
    Compute values to inject into the page
  \item
    TBD: styling
  \end{itemize}
\item
  Using a custom date picker
\end{itemize}

\hypertarget{system-implementation}{%
\section{System Implementation}\label{system-implementation}}

\emph{Goal of this section: briefly explain the current implementation.
Enough detail to ground further discussion, but don't dwell on it. It's
not the main point, and will evolve a lot.}

\begin{itemize}
\tightlist
\item
  Built as a Greasemonkey script for now (\emph{todo: convert to a full
  broser extension?})
\item
  Describe the adapter API

  \begin{itemize}
  \tightlist
  \item
    programmer creates adapters for now
  \item
    maybe end users can create in the future, or automated
  \item
    show a snippet of adapter code
  \end{itemize}
\item
  Mention the technique of scraping data from AJAX requests
\item
  future implementation goals

  \begin{itemize}
  \tightlist
  \item
    make it easy for programmers to add adapters + plugins, and
    distribute them to users. (Currently all adapters + plugins are part
    of the main Wildcard codebase)
  \end{itemize}
\end{itemize}

@@

Below are some of the ideas behind the design of Wildcard. We hope that
these principles are also useful for thinking more broadly about
enabling end users to modify software.

\hypertarget{expose-unified-structure-across-applications}{%
\subsection{Expose unified structure across
applications}\label{expose-unified-structure-across-applications}}

\emph{Todo: add a diagram here?}

In \emph{Changing Minds} \autocite{disessa2000}, Andrea diSessa
critiques the design of modern software with a story about a
hypothetical ``nightmare bike.'' Each gear on the nightmare bike is
labeled not with a number, but with an icon describing its intended use:
smooth pavement uphill, smooth pavement downhill, gravel, etc. By some
logic, this is more ``user-friendly'' than numbered gears, but in fact,
hiding orderly structure from the user makes it more difficult to
operate the bike. Many modern software designs fall into this trap,
teaching users to use isolated modes rather than coherent structure, and
the problem gets far worse when operating across multiple applications.
Unlike the UNIX philosophy of small tools interoperating through shared
abstractions, in modern computing each application is in its own silo of
data and functionality.

Wildcard helps people understand and modify the behavior of applications
through the lens of a consistent abstraction: a data table. This
abstraction strikes a balance between being both simple and generic. A
data table is simpler than the DOM tree that describes the details of
the UI, but is also generic enough to describe the essence of many
different kinds of applications.

Exposing a unified higher-level abstraction is a deliberate choice, and
is not the only way to enable users to program without directly using
the DOM. Chickenfoot \autocite{bolin2005} and CoScripter
\autocite{leshed2008} allow users to create scripts that resemble
natural language and ``sloppily'' match queries to elements in the DOM.
These designs allow for a wide range of operations, but don't explicitly
indicate what operations are possible---the user must look at the
original page and imagine the possibilities. In contrast, Wildcard
provides affordances that clearly suggest the availability of certain
types of modifications. In addition to giving users more certainty about
whether a modification is possible, these affordances might give users
new ideas for things to try. Because people are not used to modifying
web applications, providing inspiration is an important goal. (Todo:
Perhaps something to cite here, re: discoverability in GUIs vs CLIs?)

\hypertarget{direct-manipulation-by-proxy}{%
\subsection{Direct manipulation by
proxy}\label{direct-manipulation-by-proxy}}

In Wildcard, users manipulate a data table added to the page, rather
than the original page. Although the interaction with the table is
direct like using a spreadsheet, the interaction with the page is
indirectly mediated through the table.

We considered other approaches where the user would interact more
directly with the original UI, e.g.~injecting sort controls into the
page, but decided that the consistency and affordances of the table view
outweighed the potential confusion of a new layer of indirection.

Making this design successful requires maintaining a close mapping in
the user's mind between the new representation and the original page
(cite Norman? Cognitive Dimensions?). Wildcard provides live visual cues
as the user navigates the data table, similar to the highlighting
provided by browser developer tools to indicate the mapping between HTML
and the original page.

\begin{figure}
\hypertarget{fig:devtools}{%
\centering
\includegraphics{media/devtools.png}
\caption{The Chrome Dev Tools use highlighting to show the mapping
between HTML code and the page}\label{fig:devtools}
}
\end{figure}

\hypertarget{decouple-ui-from-data}{%
\subsection{Decouple UI from data}\label{decouple-ui-from-data}}

Most software does not provide much user choice for interface elements,
even for common data types: e.g., the website provides a datepicker
widget, and you have no ability to provide your own replacement
datepicker. This forces users to learn many different interfaces for
similar tasks, some of them worse than others. It also prevents
combining data across multiple applications within a single user
interface.

Wildcard exposes the data underlying an application's UI, allowing the
user to use interfaces of their choice to view and modify the data. The
Expedia datepicker demo above showed one example of how this can be
useful, but we also envision creating other widgets for visualizing and
editing data. Some examples would be showing geographic data in a custom
map that includes the user's own annotations, or editing a blog post in
a rich text editor of the user's choice.

One benefit of decoupling data from interfaces is improved UI quality.
When UI widgets can compete on their merits rather than based on network
effects from the data they have access to, it creates much stronger
competition at the interface level. For example, there is competition
among email clients (which consume an open protocol), but not among
Facebook or Twitter clients. This benefit relates to the SOLID project
led by Tim Berners-Lee \autocite{berners-lee2018}, which envisions
user-controlled data as a mechanism for decoupling data from interfaces,
e.g.~giving users a choice of which client to use to consume a given
social media feed. Wildcard has overlapping goals with SOLID, but does
not require decentralized user control of data---the data can remain on
a centralized server, as long as the interface can be tweaked by end
users. \emph{Note: maybe this paragraph belongs in related work?}

There is also value in using the same consistent interface across many
applications. For example, many programmers become deeply familiar with
one text editor and use it for many different kinds of tasks. This usage
even extends beyond editing permanent files; shell programs can use the
user's preferred text editor as an interactive input mechanism (e.g.~for
editing git commit messages). This ability to generically reuse the text
editor in many contexts makes it more worthwhile to invest time
mastering the tool. Beaudouin-Lafon and Mackay refer to this type of
reuse as \emph{polymorphism} in interaction
\autocite{beaudouin-lafon2000}, noting that it is a key technique for
keeping interfaces simple while increasing their power.

\emph{Note: Maybe could relate this section to Concept Reuse?}

\emph{Note: Does this section overlap with ``expose unified
structure''?}

\hypertarget{in-place-toolchain}{%
\subsection{In-place toolchain}\label{in-place-toolchain}}

\begin{longtable}[]{@{}lll@{}}
\toprule
\endhead
& \textbf{In-place} & \textbf{Not in-place}\tabularnewline
\textbf{End user friendly} & Wildcard & IFTTT\tabularnewline
\textbf{For developers} & browser dev tools & editing OSS\tabularnewline
\bottomrule
\end{longtable}

We believe that people are more motivated to modify software when the
tools for modification are close at hand while using the original
software, referred to by Ink and Switch as an ``in-place toolchain''
\autocite{inkandswitch2019}.

Browser developer tools exemplify an in-place toolchain, since they are
built in to browsers and available in the context of viewing a page; in
contrast, creating a browser extension requires moving out of the
browser to a separate development environment.

Wildcard creates an in-place toolchain for modifying web applications.
Once installed, if an adapter is available for a site, it can be
immediately modified from within the browser. Hopefully, this can help
make switching from using to modifying software into a smooth
transition, rather than a chasm to cross.

\hypertarget{first-party-cooperation-optional}{%
\subsection{First party cooperation
optional}\label{first-party-cooperation-optional}}

The Web is an unusually extensible platform. On many other platforms
(e.g.~smartphone operating systems), software is locked down unless
first-party developers explicitly provide hooks for plugins and
interoperability, but on the Web, all client-side code is available for
browser extensions to modify. Application authors can use practices that
make it easier to modify their apps (e.g.~clean semantic markup), or
more difficult (e.g.~code obfuscation), but the default state is
openness. This gives extensions freedom to modify applications in
creative ways that the original developers did not plan for.

Wildcard takes advantage of this openness, and does not depend on
cooperation from first-party website developers. Any programmer can add
support for any website to Wildcard by building a third party adapter.
This design acknowledges the pragmatic reality that most sites today
will not have Wildcard support built in, and more broadly do not
prioritize client side extensibility.

However, we also hope that first party website developers eventually
build in Wildcard support to their applications, which would reduce the
burden of maintaining adapters and make Wildcard plugins more stable.
While this might seem optimistic, implementing the Wildcard adapter API
could be straightforward in a typical client-side application that
already has access to a structured version of the data in the page.
There is also precedent for first parties implementing an official
client extension API in response to user demand: for several years,
Google exposed an extension API in Gmail for Greasemonkey scripts to
use. (Since then, third parties have created well-maintained
alternatives \autocite{streak,talwar2019}, illustrating the value of
sharing even third party adapters.)

>>>>

\author{Geoffrey Litt}
\authorinfo{is a PhD student at MIT, researching programming tools for end users and developers.}
\author{Daniel Jackson}
\authorinfo{is a professor in the Department of Electrical Engineering and Computer Science at MIT, associate director of CSAIL, and a MacVicar Fellow. He leads the Software Design Group.}
\affiliation{Massachusetts Institute of Technology}

\keywords{end-user programming, software customization, web browser extensions} % please provide 1--5 keywords

%%%%%%%%%%%%%%%%%%
%% These data MUST be filled for your submission. (see 5.3)
\paperdetails{
  %% perspective options are: art, sciencetheoretical, scienceempirical, engineering.
  %% Choose exactly the one that best describes this work. (see 2.1)
  perspective=art,
  %% State one or more areas, separated by a comma. (see 2.2)
  %% Please see list of areas in http://programming-journal.org/cfp/
  %% The list is open-ended, so use other areas if yours is/are not listed.
  area={Programming environments, Visual and live programming},
  %% You may choose the license for your paper (see 3.)
  %% License options include: cc-by (default), cc-by-nc
  % license=cc-by-sa,
}
%%%%%%%%%%%%%%%%%%

@@

% Please always include the abstract.
% The abstract MUST be written according to the directives stated in 
% http://programming-journal.org/submission/
% Failure to adhere to the abstract directives may result in the paper
% being returned to the authors.
\begin{abstract}
Many Web applications do not meet the precise needs of their users.
Browser extensions and user scripts offer a way to customize web
applications, but most people do not have the programming skills to
implement their own customizations.

In this paper, we present a prototype of Wildcard, a live programming
system embedded in the browser that empowers users to casually tweak web
applications without needing to program in Javascript or interact with
the DOM.

Wildcard shows the data from a web page in a table, and maintains a
bidirectional connection between the table and the original page. By
directly manipulating the table, users can perform a wide variety of
modifications: sorting/filtering content, adding private annotations,
using spreadsheet formulas to fetch data from other web services, using
custom UI elements to edit form data, and more.

We present examples of using Wildcard to solve real world problems, and
explain the design principles behind the prototype. In the future, we
envision continuing to build Wildcard into a fully deployed system that
helps make the web into a more malleable medium for all users.
\end{abstract}

@@

In 2012, the travel site Airbnb removed the ability to sort listings by
price. Users could still filter by price range, but could no longer view
the cheapest listings first. Many users complained on online message
boards that the change seemed hostile to users. ``It's so
frustrating!..What is the logic behind not having this function?'' said
one user on the
\href{https://community.withairbnb.com/t5/Hosting/Sorting-listing-by-price/td-p/559404}{Airbnb
support forum}. Alas, the feature remains missing to this day.

This is a familiar situation in a world of web applications that are
frequently updated without user consent. Sometimes there is a browser
extension or user script that fixes an issue, and if the user is a
motivated, programmer they might even be able to implement their own
fix. But for most people in most situations, the only recourse is to
complain to the developers and pray that someone listens---or more
likely, to simply give up. While many have become used to this status
quo, we see it as a tremendous waste of the openness of the Web
platform. In \emph{Personal Dynamic Media} \autocite{kay1977}, Alan Kay
originally envisioned personal computing as a medium that let a user
``mold and channel its power to his own needs,'' but today's software is
more like concrete than clay.

In this paper, we introduce Wildcard, a live programming system embedded
in the browser that enables users to tweak web applications without
programming in Javascript or interacting with the DOM. Wildcard adds a
panel to the bottom of a web page that shows a structured table view of
the main data in the page. The table maintains a bidirectional
connection to the original page---when the user manipulates the table,
the original page gets modified, and vice versa.

In Wildcard, a user can sort Airbnb listings with just one intuitive
click on a table header, with no programming required. Beyond sorting
and filtering data, Wildcard also supports accessing third party APIs,
performing small computations, recording private user annotations, and
using alternate UI widgets to control a page. While Wildcard does not
support any change someone might want to make to a website, it makes a
variety of common and useful changes more accessible to end users, as
well as to programmers who want to avoid tedious low-level programming.

Under the hood, the implementation of Wildcard is straightforward,
because a programmer must manually write an adapter for each website
which uses common web scraping techniques to map the data from the web
page into the Wildcard table format. Although there is still manual
programming involved, the result is very different from traditional
single-use browser extensions, because instead of the programmer
defining a narrow use case, the end user is able to make many different
changes on top of a single site-specific adapter.

In this paper, we present examples of using Wildcard to solve real world
problems, and then explain the system architecture and design principles
behind the prototype:

\begin{itemize}
\tightlist
\item
  \emph{Expose consistent structure}: turning isolated applications into
  a structured medium
\item
  \emph{Encourage casual tweaking}: using frequent software modification
  as a starting point for end user programming
\item
  \emph{Build for an ecosystem}: designing for first party developers,
  third party programmers, and non-programmer end users
\end{itemize}

We also explore related work on web augmentation tools,
spreadsheet-based app builders, and web scraping tools. Our novel
contribution in this paper is to combine ideas from each of these areas
into a new kind of tool: a programming environment which maps data from
web pages onto a live table view, enabling end users to modify the data.

Wildcard is currently an early research prototype, with incomplete
features and limited coverage of sites. We plan to continue privately
testing the system with our own use cases, and then to release the tool
publicly, enabling other programmers to contribute site adapters and
other plugins. Ultimately, we hope that Wildcard makes modifying
websites into a natural part of using the web, as mundane and ubiquitous
as using spreadsheets.

\hypertarget{demos}{%
\section{Demos}\label{demos}}

To get a sense of how it feels to use Wildcard, let's see an example of
using it to help with booking a trip using the travel sites Airbnb and
Expedia.

\hypertarget{sorting-and-filtering}{%
\subsection{Sorting and filtering}\label{sorting-and-filtering}}

We start by opening up the Airbnb search listings page to look for a
place to stay. As mentioned before, this page doesn't let us sort by
price, but we can use Wildcard to fix that. First, we open up the
Wildcard panel, which shows a table corresponding to the search results
in the page. As we click around in the table, the corresponding row in
the page is highlighted so we can see the connection between the views.

When we click the Price column header in the table, both the table and
the original page become sorted by price. We can also filter listings by
rating, another feature not offered in the Airbnb site.

After finishing the sorting and filtering, we can close the table view
and continue using the website in its original polished design.

\hypertarget{row-actions}{%
\subsection{Row actions}\label{row-actions}}

Most websites that show lists of data also offer actions that can be
taken on a row in the table, like adding an item to a shopping cart.
Wildcard has the ability to make these actions available in the data
table if the site adapter implements them. The main advantage this
provides is the ability to easily perform an action in bulk across
multiple rows.

For example, it's tedious on Airbnb to click on listings one by one to
add them to a list of favorites. Using Wildcard, we can just select
multiple rows and favorite all of them with one click. Similarly, we can
also open the detailed pages for many listings in new tabs.

Within the site adapter, each action is implemented as a regular
Javascript function running in the context of the page, which can click
on buttons in the UI, launch AJAX requests, navigate to a new page, etc.

\hypertarget{user-annotations}{%
\subsection{User annotations}\label{user-annotations}}

It's a common practice in spreadsheets to add a column next to a table
for recording notes about the item in each row. In Wildcard, this
pattern can be used to create private annotations in a website. Here, we
use this feature to jot down pros and cons of various listings:

In order to match the annotations with the existing design, the site
adapter specifies a reasonable location in each row for injecting new
data.

Annotations are persisted in browser local storage so the user can come
back and view them later. The site adapter associates each table row
with a stable identifier from the original site so that annotations can
be displayed with the appropriate item.

Currently annotations are private to the browser of the user that
created them, but it could be useful in the future to allow users to
share annotations.

\hypertarget{computation-with-formulas}{%
\subsection{Computation with formulas}\label{computation-with-formulas}}

The demos so far have shown straightforward tweaks that provide useful
conveniences with little effort. But Wildcard also supports more
sophisticated tweaks that fetch external data and perform computations,
using a formula language.

When traveling without a car, it's nice to evaluate potential places to
stay based on how walkable the surroundings. Using Wildcard formulas, we
can integrate Airbnb with Walkscore, an API that can rate the
walkability of any location on a 1-100 scale. When we call the
\texttt{WALKSCORE} formula with the latitude and longitude of the
listing, it returns the score as the cell value. Because the cell is
shown in the page, the score also shows up in the page body.

Wildcard formulas are just Javascript functions that take table data as
input and return new data. Formulas can have limited side effects like
accessing external APIs to fetch data, but do not directly manipulate
the page. Instead, they return data into the table, which is then
injected into the page with the same mechanism used for injecting user
annotations.

\hypertarget{custom-ui-elements}{%
\subsection{Custom UI elements}\label{custom-ui-elements}}

It might seem that Wildcard is only useful on websites that display
lists of tabular data like search results. But in fact, the table
metaphor is flexible enough to represent many types of data. For
example, a form can be represented as a single row, with a column for
each input.

In previous examples the data extracted from the site was marked as
read-only (it's not allowed to change the name of an Airbnb listing),
but in this case, the cells are marked as writable, so that changes in
the table are reflected in the original form. This becomes particularly
useful when combined with custom cell editing UIs, which allow users to
graphically edit the value of a cell.

Here's an example of using those features to help with filling in a
form. Filling in dates for a flight search typically requires opening up
a separate calendar app to look up the correct dates, and then manually
copying them into the form. In Wildcard, we can make this easier by
editing the date cell using a datepicker widget that has privileged
access to our calendar information.

Custom UIs enable people to use a consistent UI to enter common types of
data across the web. They also allow a user to access their own private
data as part of a web interface, without needing to expose it to the
website server.

Overall, an interactive data table is a computation model that presents
a surprisingly large range of useful possibilities for end user
modification of websites, while also remaining familiar and easy to use.
While we've presented concretely illustrated here just a few of these
possibilities, we plan to continue developing site adapters, formulas,
and UI elements to explore more use cases, and to eventually publicly
release the tool to allow real end users to discover their own uses.

\hypertarget{system-architecture}{%
\section{System Architecture}\label{system-architecture}}

Wildcard is written in Typescript. It is currently injected into pages
using the \href{https://www.tampermonkey.net/}{Tampermonkey} userscript
manager, but in the future we plan to deploy it as a standalone browser
extension to make it easier to install.

In order to maximize extensibility, Wildcard is implemented as a small
core program along with several types of plugins: site adapters,
formulas, and cell renderers/editors. The core contains functionality
for displaying the data table and handling user interactions, and the
table implementation is built using the
\href{https://handsontable.com/}{Handsontable} Javascript library.

\begin{figure}
\centering
\includegraphics{media/architecture-clean.png}
\caption{The architecture of the Wildcard system}
\end{figure}

Site adapters specify the bidirectional connection between the web page
and its structured data representation.

For extracting data from the page and getting it into structured form,
Wildcard provides ways to concisely and declaratively express scraping
logic. For example, here is a code snippet for extracting the name of an
Airbnb listing. It specifies the field name and some metadata
properties, and then a function that extracts the name div from a row
div.

\begin{Shaded}
\begin{Highlighting}[]
  \OperatorTok{\{}
\NormalTok{    fieldName}\OperatorTok{:} \StringTok{"name"}\OperatorTok{,}
\NormalTok{    readOnly}\OperatorTok{:} \KeywordTok{true}\OperatorTok{,}
\NormalTok{    type}\OperatorTok{:} \StringTok{"text"}\OperatorTok{,}
\NormalTok{    el}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \VariableTok{row}\OperatorTok{.}\FunctionTok{querySelector}\NormalTok{(}\VerbatimStringTok{\textasciigrave{}.}\SpecialCharTok{$\{}\NormalTok{titleClass}\SpecialCharTok{\}}\VerbatimStringTok{\textasciigrave{}}\NormalTok{)}\OperatorTok{,}
  \OperatorTok{\}}
\end{Highlighting}
\end{Shaded}

The site adapter also needs to support the reverse direction: sending
updates from the table to the original page. Most DOM manipulation is
not performed directly by the site adapter: the adapter specifies how to
find the divs representing data rows, and the core platform mutates the
DOM to reflect the table state. The only exception is site-specific
actions (like favoriting in Airbnb) which runs Javascript code that can
mutate the DOM or perform other arbitrary behaviors.

So far we have only implemented three site adapters in Wildcard, but we
plan to implement many more in the near future to continue to test the
limits of the API. Eventually we will also encourage third party
programmers to build and maintain adapters for popular sites.

@@

The design of Wildcard is grounded in several principles, informed by
prior work and our own experimentation. We think these principles extend
beyond this specific system and can more generally inform the design of
end user programming tools, particularly systems enabling users to
modify GUI apps.

\hypertarget{expose-consistent-structure}{%
\subsection{Expose consistent
structure}\label{expose-consistent-structure}}

In \emph{Changing Minds} \autocite{disessa2000}, Andrea diSessa
critiques the design of modern software with a story about a
hypothetical ``nightmare bike.'' Each gear on the nightmare bike is
labeled not with a number, but with an icon describing its intended use:
smooth pavement uphill, smooth pavement downhill, gravel, etc. This
might seem more ``user-friendly'' than numbered gears, but in fact, it
makes it harder to operate the bike. An ordered numerical sequence
allows the user to intuitively grasp the structure of the system---``if
I shift up, it gets harder to pedal''---but isolated modes provide a
superficial understanding with no grounding in structure. This
understanding might be sufficient for the most common cases, but breaks
down in unfamiliar situations. If someone needs to go uphill on gravel,
do they need to try every mode at random?

Today, most personal computing consists of using distinct apps, each one
an isolated silo of concepts, interfaces, data, and functionality. While
there are some shared idioms and limited points of interoperability, the
fundamental expectation is that each app operates independently. If a
new need emerges, a user must hope ``there's an app for that''; creating
a new tool to meet the need is often out of the question.

Computing does not need to operate this way. As one example, UNIX offers
a compelling alternate vision of small tools connected through a common
abstraction of text files. Just like riding a bike with numbered gears,
UNIX requires users to build up a structured understanding of the
system, which then allows them to flexibly react to new needs not served
by existing tools. Users can also deeply master a text editor and use it
for a vast array of tasks, even as an interactive input mechanism in
shell programs (e.g.~for editing git commit messages). Beaudouin-Lafon
and Mackay call this style of interaction \emph{polymorphic}
\autocite{beaudouin-lafon2000}, since a single interface can be used in
many contexts. diSessa \autocite{disessa2000} notes the deep connection
between polymorphism and literacy: textual literacy rests on a single
rich medium of writing, which once mastered, can be adapted to many
different genres and uses. If people needed to relearn from scratch when
switching from writing essays to writing emails, the medium would lose
most of its potency; this is the situation in software today.

With Wildcard, we hope to make isolated Web applications work a little
bit more like UNIX, by imposing a consistent and simple structure onto
many isolated applications. Just as UNIX uses text files to represent
all data, Wildcard uses relational tables. This abstraction strikes a
balance between being simple and generic: a data table is much simpler
than the DOM tree that describes all the details of the UI, but is also
generic enough to describe the essence of many different interfaces.

\begin{figure}
\centering
\includegraphics{media/structure.png}
\caption{Wildcard layers the consistent structure of a data table over
different isolated applications.}
\end{figure}

Exposing more structured data to users provides improved quality and
consistency of user interfaces. When UI widgets can compete on their
merits rather than having a monopoly on a given data silo or
application, it creates stronger competition at the interface level. For
example, there is competition among email clients which consume an open
protocol, but not among Facebook or Twitter clients. Exposing structure
also leads to improved UI consistency across apps: UNIX users can reuse
their text editor everywhere, but Web users must use a different rich
text editor for each site. Tim Berners-Lee has written in more detail
about these benefits \autocite{berners-lee2018} in the context of the
SOLID project, which envisions user-controlled data as a mechanism for
decoupling data from interfaces, e.g.~giving users a choice of which
client to use to consume a given social media feed. Wildcard has
overlapping goals with SOLID, but does not require decentralized user
control of data---the data can remain on a centralized server, as long
as the interface can be tweaked by end users.

Layering a structured view on top of an existing application is not the
only option available. Systems like Chickenfoot \autocite{bolin2005} and
CoScripter \autocite{leshed2008} forego the intermediate layer of
structure, instead allowing users to create scripts in an informal
language and then perform fuzzy pattern matching to find elements in the
DOM. For example, to find a button after a textbox in Chickenfoot, the
user could type \texttt{click(find(“button\ just\ after\ textbox”))}.
These designs allow for expressing a wide range of operations, but they
don't explicitly indicate what operations are possible---the user can
only see the original page and imagine the possibilities. In contrast,
Wildcard provides affordances that clearly suggest the availability of
certain actions (e.g.~sorting, editing a cell, adding a column with a
derived value), especially to users who are familiar with spreadsheets.
In addition to giving users more certainty about whether a modification
is possible, these affordances might improve discoverability and give
users new ideas for things to try. We also think that sloppy programming
does little to solve the problem of thinking of applications as isolated
worlds, whereas Wildcard contributes to giving users a structured
understanding of the different apps that they use.

One key challenge of making a structured view successful is ensuring a
close mapping in the user's mind between the new representation and the
original page. Wildcard provides live visual cues as the user navigates
the data table (similar to the highlighting provided by DOM inspectors
in browser developer tools). In our own usage, we have found that this
live highlighting makes it very clear how the two representations map to
each other.

\hypertarget{encourage-casual-tweaking}{%
\subsection{Encourage casual tweaking}\label{encourage-casual-tweaking}}

In order to serve the widest possible audience, Wildcard is designed for
tweaking existing software for one's own needs, not for designing a new
application from scratch. Creating a new application requires motivation
and other skills besides programming that not everyone has, but having
frustrations with some aspect of an app which could be solved by a small
tweak is a nearly universal experience.

\begin{longtable}[]{@{}lll@{}}
\toprule
\endhead
& \emph{Casual} & \emph{Not casual}\tabularnewline
\emph{End user friendly} & \textbf{Wildcard} & IFTTT\tabularnewline
\emph{Requires programming} & browser dev tools & editing open source
software\tabularnewline
\bottomrule
\end{longtable}

We can evaluate a system for tweaking software along two dimensions.
First, the technical capability required of the user: is programming
knowledge needed? Second, the level of effort required: how far out of
their way the user must go to make a change? Can they casually make a
small change, or do they need to make a larger project out of it? These
dimensions are not orthogonal, but they are distinct. For example,
setting up a workflow trigger in an end user programming system like
\href{https://ifttt.com/}{IFTTT} does not require much technical skill,
but it does require leaving the user's normal software and entering a
separate environment. On the other hand, running a Javascript snippet in
the browser console requires programming skills, but can be done
immediately and casually in the flow of using a website.

In addition to requiring no programming skills, Wildcard also aims to
support frequent, small modifications. The Wildcard table appears in the
course of normal web browsing, to ensure that the tools for modification
are close at hand while using the original software. Ink and Switch
refers to this property as having an ``in-place toolchain''
\autocite{inkandswitch2019}. Wildcard also provides live feedback,
immediately editing the page in response to user interaction, which
gives users confidence in the changes they are making.

Wildcard also aims to make some valuable changes possible with
particularly low effort: for example, sorting a table in a single click,
or adding an annotation by filling in a cell. We would expect many
Wildcard users to start by using these bits of functionality, before
possibly moving on to more complex features like formulas. This property
is inspired by spreadsheets, which can be used productively even by
someone who has learned only a tiny part of their functionality,
e.g.~for storing tables of numbers or computing simple sums
\autocite{nardi1991}. In contrast, many traditional programming systems
require someone to spend days learning complex concepts before being
able to create a useful piece of software---printing ``hello world'' is
not a genuinely valuable activity.

\hypertarget{build-for-an-ecosystem}{%
\subsection{Build for an ecosystem}\label{build-for-an-ecosystem}}

Some end user programming systems are designed with a single user in
mind---a lone non-programmer, needing to achieve a task through some
combination of individual skills and a powerful tool. But in fact,
real-world usage of end user programming systems is far more
collaborative. Spreadsheets are built together by users with different
levels of technical expertise: some users just write simple formulas,
while their officemates help write more complex formulas or even program
macros in traditional languages \autocite{nardi1990}. Our goal with
Wildcard is to consider the full ecosystem of users and to design a tool
that best combines all of their abilities.

This design principle is most salient in our approach to building site
adapters. Many web augmentation projects aim to make the full process of
tweaking available to end users with no programmer help needed, through
a combination of automated heuristics and interactive feedback. This is
a laudable goal because it makes these tools work with any site, and
these approaches are often quite successful, but they also inevitably
break down sometimes \autocite{bolin2005} \autocite{little2010}. In
contrast, human programmers working together have been able to maintain
stable scraping libraries, even for complex sites like Gmail
\autocite{streak,talwar2019}. Our primary approach to building site
adapters is having programmers manually create them, because we are
willing to trade off lower site coverage in return for more reliable
results on the sites that are covered, building confidence for users.
Like the collaboration between an expert and novice spreadsheet user,
this architecture leverages more skilled programmers while still
providing end users with a flexible programming environment to meet
their own needs. On the other hand, we remain very open to incorporating
other tools in the future that enable end users to also create their own
adapters.

\begin{figure}
\centering
\includegraphics{media/adapters.png}
\caption{Having third party programmers create site adapters balances
site coverage and adapter quality.}
\end{figure}

In any system for tweaking software, another important consideration is
how to involve the first party developers of the original software. Our
approach is to avoid depending on their cooperation, but also to aim for
eventual cooperation.

This approach takes advantage of the unusual openness of the Web
platform. On many other platforms (e.g.~smartphone operating systems),
software is locked down unless first-party developers explicitly provide
hooks for plugins and interoperability, but on the Web, all client-side
code is available for browser extensions to modify. Application authors
can use practices that make it easier to modify their apps (e.g.~clean
semantic markup), or more difficult (e.g.~code obfuscation), but the
default state is openness. This gives extensions freedom to modify
applications in creative ways that the original developers did not plan
for. At the most extreme, this can even include modifications that are
hostile to the desires of the original developers, a notion Cory
Doctorow calls
\href{https://www.eff.org/deeplinks/2019/10/adversarial-interoperability/}{adversarial
interoperability}. Although ad blocking extensions make this style of
extension very prominent, most ideas for extending software merely fall
outside the range of possibilities originally imagined by the developer,
rather than being actively hostile.
